An upper bound to gas storage and delivery via pressure-swing adsorption in porous materials

### 2021-07-09 gas-adsorption/rsc-gas-ads-complete.tex
k

We follow Coudert et al. \cite{coudert2008thermodynamics} in developing a
thermodynamic model for a two-phase adsorbent. Let $\Delta
F=F_{\text{open}}-F_{\text{closed}}$ be the Helmholtz free energy difference per
mole of adsorbent between the two phases in the absence of gas. Suppose the
closed phase is more stable ($\implies \Delta F >0$), so that upon a reduction
in the pressure of the gas, the material will collapse and expel all residual
gas to enhance the deliverable capacity compared to rigid materials that will
retain some gas at $\pempty$. In addition, assume the open phase hosts a
uniform, attractive potential energy field $\V<0$ for gas molecules to adsorb,
so that eqn.~\ref{eq:xi_vs_xi0} holds for the open phase of the material.  This
approximate model of the two-phase material has no upper bound on the
deliverable capacity, since $\V$ and $\Delta F$ can together be increased
without bound.  We can, however, place a bound for a fixed $\Delta F$, since if
the gas is too strongly attracted to the open phase, the material will remain
open at $\pempty$ and our rigid upper bound applies.

The upper bound for the two-phase material is found by identifying the
attractive interaction $\V$ that balances the gas-host attractions in the
idealized-open-substrate at $\pempty$ with the free energy penalty to open.

In
terms of $\rho_g(\mu, T)$, then, the grand potential $\Omega$ in our idealized
substrate is:
\begin{equation}
    \Omega(T, \mu) = - \int_{-\infty}^\mu \rho_g(\mu-\V)d\mu.
\end{equation}
However, the \emph{osmotic potential} governs which phase the material adopts.
The material changes phases when osmotic potentials of the two phases are equal.
% At the pressure at which they switch when the osmotic potentials in the two phases
% are equal.
The upper bound for the two-phase material is found by identifying the largest
possible attraction $|\V|$ consistent with the material transitioning between
$\pempty$ and $\pfull$ to expel all residual gas.
%
This means the material should change phases at pressure $\pempty$, which gives
\begin{align}
    % F_{closed} + 0 &= F_{open} + \Omega(T, \mu) \\
    \frac{\Delta F}{V_{\text{open}}} &= \int_{-\infty}^{\muempty} \rho_g(\mu-\V_{opt})d\mu.
\end{align}
where we follow Coudert \emph{et al.}\cite{coudert2008thermodynamics} and
neglect the $pV$ terms, and $V_{\text{open}}$ is the molar volume of the open
phase.

% David: introduce units of delta G here.

For an enhancement of the deliverable capacity by this two-phase, flexible material, 
the open phase must be preferred at $\pfull$, with the closed phase preferred at $\pempty$. 
This requires that, neglecting gas-gas interactions:
% David: doesn't this also neglect entropy? wut about the *free energy* of the gas?
% David: I suggest \Delta G_{host} to be clear?
\begin{align}
    V_{\text{open}}|\rho_{\text{open}}(\pfull)\V| &> \Delta F \\
    V_{\text{open}}|\rho_{\text{open}}(\pempty)\V| &< \Delta F
\end{align}
% David: I put \V<0, think we should say the sign of \Delta G to be clear too.
% David: specify units on \delta G so clear where V comes from; might be better to absorb V into delta G?
where $\rho_{\text{open}}(p)$ is the density of gas
in the open phase of the material at pressure $p$ and $V_{\text{open}}$ is the molar volume of the open phase of the adsorbent, and we neglect the $pV$
terms in the Gibbs free energy.
%under the assumption that the mechanical
%work associated with the transition is neglible when compared with the
%adsorption energies involved.
I.e., (1, to be open at $\pfull$) the gas-host attractions at $\pfull$ must
overcome the free energy penalty to open the structure and (2, to be closed at
$\pempty$) the free energy penalty to open the structure must be larger than the
(weak) gas-host attractions at $\pempty$. This approximate model of the
two-phase material has no upper bound on the deliverable capacity, since $\V$
and $\Delta F$ can together be increased without bound. We can, however, place a
bound for a fixed $\Delta F$, since if the gas is too strongly attracted to the
open phase, the material will remain open at $\pempty$ and our rigid upper bound
applies.
% David: does our rigid upper bound apply then?
The upper bound for the two-phase material is found by identifying the
attractive interaction $\V$ that balances the gas-host attractions in the
idealized-open-substrate at $\pempty$ with the free energy penalty to open:
\begin{align}
    V_{\text{open}} \rho_g(\muempty{} - \V_{opt}, T)\V_{opt} = \Delta F.
\end{align}
% David: left is negative. what is the right? negative as well? how did u define it?
This gives the largest possible attraction $|\V|$ consistent with the material closing
at $\pempty$ to expel all residual gas. 
The deliverable capacity is then the density of gas in the open phase at $\pfull$, $\rho_g(\mufull{} - \V_{opt}, T)$.
% David: changed it to \V_opt, or not necessary?

Figure~\ref{fig:methane-flexible} illustrates
how the upper bound on the deliverable capacity of methane in a two-phase,
collapsing material depends on $\Delta F$. 
For comparison, we include the free energy
difference for Co(bdp)~\cite{choi2008broadly} and its measured deliverable capacity.  
While this upper bound is not rigorous, it suggests that a two-phase flexible material for methane storage
would benefit from significantly greater stability of the closed phase---provided the open phase
had correspondingly greater attraction for the gas molecules.  Figure~\ref{fig:hydrogen-flexible}
shows the corresponding result for hydrogen storage, indicating that, for a flexible material to
reach the DOE ULTIMATE deliverable capacity, it would require far greater stability for the closed
phase, as well as an unprecedented level of attractive interaction for hydrogen.

>>>>

We follow Coudert et al. \cite{coudert2008thermodynamics} in developing a
thermodynamic model for a two-phase, collapsing adsorbent like Co(bdp). 
Let $\Delta F=F_{\text{open}}-F_{\text{closed}}$ be the Helmholtz free energy difference per
mole of adsorbent between the two phases in the absence of gas. Suppose the
closed phase is more stable ($\implies \Delta F >0$), so that upon a reduction
in the pressure of the gas, the material can collapse and expel all residual
gas to enhance the deliverable capacity compared to rigid materials that will
retain some gas at $\pempty$. In addition, assume the open phase hosts a
uniform, attractive potential energy field $\V<0$ for gas molecules to adsorb,
so that eqn.~\ref{eq:xi_vs_xi0} holds for the material when in the open phase.

This model of a two-phase, collapsing material has no upper bound on the
deliverable capacity, since $|\V|$ and $\Delta F$ can together be increased
without bound.
We can, however, place a bound on the deliverable capacity for a fixed $\Delta F$, since if
the gas is too strongly attracted to the open phase, the material will remain
open at $\pempty$, and, then, our rigid upper bound applies.
We place an upper bound on the deliverable capacity in the two-phase material by identifying the
attractive interaction $\V$ that balances the gas-host attractions in the
idealized-open-substrate at $\pempty$ with the free energy penalty to open.

For an enhancement of the deliverable capacity by this two-phase, flexible material, 
the open phase must be preferred at $\pfull$, with the closed phase preferred at $\pempty$. 
% At the pressure at which they switch when the osmotic potentials in the two phases
% are equal.
The upper bound for the two-phase material is found by identifying the largest
possible attraction $|\V|$ consistent with the material transitioning between
$\pempty$ and $\pfull$ to expel all residual gas.
The material changes phases when the osmotic potentials of the open and closed systems 
with and without gas, respectively, are equal \cite{coudert2008thermodynamics} (neglecting any hystersis).
Thus, the optimal background potential $\V_{opt}$ equates the grand potential of the gas in the open phase 
with the free energy difference between the two phases at $\pempty$:
%In terms of $\rho_g(\mu, T)$, the grand potential $\Omega$ in our idealized
%substrate is:
%\begin{equation}
%    \Omega(T, \mu) = - \int_{-\infty}^\mu \rho_g(\mu-\V)d\mu.
%\end{equation}
%
% This means the material should change phases at pressure $\pempty$, which gives
\begin{align}
    % F_{closed} + 0 &= F_{open} + \Omega(T, \mu) \\
    \frac{\Delta F}{V_{\text{open}}} &= \int_{-\infty}^{\muempty} \rho_g(\mu-\V_{opt})d\mu, \label{eq:v_opt_bistable}
\end{align}
where we follow Coudert \emph{et al.}\cite{coudert2008thermodynamics} and
neglect the $pV$ terms, and $V_{\text{open}}$ is the molar volume of the open
phase.
I.e., $|\V_{opt}|$ gives the maximal gas-open-host attractions under the constraint that the material collapses and expels all residual gas at $\pempty$; if the attraction were greater, the material would refuse to collapse and expel its gas at $\pfull$.
The deliverable capacity is then the density of gas in the open phase at $\pfull$, $\rho_g(\mufull{} - \V_{opt}, T)$ with $\V_{opt}$ satisfying eqn.~\ref{eq:v_opt_bistable}.
% David: changed it to \V_opt, or not necessary?

Figure~\ref{fig:methane-flexible} illustrates
how the upper bound on the deliverable capacity of methane in a two-phase,
collapsing material depends on $\Delta F$-- and the $\V_{opt}$ that maximizes the deliverable capacity. 
For comparison, we include the free energy
difference for Co(bdp)~\cite{choi2008broadly} and its measured deliverable capacity.
Indeed, the deliverable capacity of a two-phase, collapsing material can exceed that of a rigid material if 
(i) the closed phase is sufficiently stable relative to the open phase
and
(ii) the gas has a correspondingly high affinity for the open phase.
Figure~\ref{fig:hydrogen-flexible} shows the corresponding result for hydrogen storage, indicating that, for a flexible material to
reach the DOE ULTIMATE deliverable capacity, it would require far greater stability for the closed
phase, as well as an unprecedented level of attractive interaction for hydrogen.
